taxadb: A High-Performance Local Taxonomic Database Interface

### 2020-07-20 paper/paper.Rmd
:construction: boilerplate

---
output: 
  github_document:
    df_print: tibble
---

@@

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = TRUE
)
library(neonstore)
Sys.setenv("NEONSTORE_HOME" = tempfile())
```

# neonstore

<!-- badges: start -->
[![R build status](https://github.com/cboettig/neonstore/workflows/R-CMD-check/badge.svg)](https://github.com/cboettig/neonstore/actions)
[![Codecov test coverage](https://codecov.io/gh/cboettig/neonstore/branch/master/graph/badge.svg)](https://codecov.io/gh/cboettig/neonstore?branch=master)
<!-- badges: end -->

`neonstore` provides quick access and persistent storage of NEON data tables. 
`neonstore` emphasizes simplicity and a clean data provenance trail, see 
(Provenance)[/#Provenance].

@@

```{r}
neon_citation()
```

@@

>>>>

---
title: "taxadb: A High-Performance Local Taxonomic Database Interface"
author:
  - name: "Carl Boettiger"
    affiliation: ucb
  - name: "Christine Laney"
    affiliation: neon
  - name: "Quinn Thomas"
    affiliation: vt
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: neon
    address: "NEON"
  - code: vt
    address: "Virginia Tech"
abstract: |
  The National Ecological Observatory Network
journal: "TBD"
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: rticles::elsevier_article
---

@@

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = TRUE
)
library(neonstore)
Sys.setenv("NEONSTORE_HOME" = tempfile())
```

The National Ecological Observatory Network (NEON) provides open access ecological data as 181 (active and planned) data products at 81 (permanent and relocatable) sites throughout the United States.  The `neonstore` R package seeks to provide quick, convenient access and persistent storage of NEON data files.  `neonstore` design emphasizes simplicity and data provenance. here, we describe the design, use, and rationale for the package.

NEON data products are built around a highly atomized file design: both observation systems (OS) and instrument systems (IS) data products are divided into separate files by site and month.  Products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.  This atomized approach facilitates sub-setting and data transfer, (particularly relevant to R users where most objects are typically stored in memory)

@@

```{r}
## Note: NEON's recommendation may soon change
neon_citation()
```

@@

## Non-stacking files and low-level interface

At it's core, `neonstore` is simply a mechanism to download files from the NEON API.
While the `.csv` files from the Observation Systems (OS, e.g. bird count surveys),
and Instrument Systems (e.g. aquatic sensors) are typically stacked into large
tables, other products, such as the `.laz` and `.tif` images produced by the
airborne observation platform LIDAR and cameras may require a different approach.

```{r}
# Read in a large file list for illustration purposes
cper_data <- readr::read_csv("https://minio.thelio.carlboettiger.info/shared-data/neon_data_catalog.csv.gz")

## Typically one would read all files in local store, e.g. list.file(neon_dir())
df <- neon_filename_parser(cper_data$name)
```

```{r}
library(dplyr)
df %>% count(EXT, sort=TRUE)
```

We can take a look at all `laz` LIDAR files:

```{r}
df %>% 
  filter(EXT == "laz")
```

Note that many of the airborne observation platform (AOP) products, such as
these LIDAR files, do not include the PRNUM or REV components that make up part
of the `productCode`s used in the NEON `product` tables.
